---
title: "Multivariate analysis of Gaussian data (MANOVA)"
subtitle: Part I
author: |
  | Wagner H. Bonat   | Walmes M. Zeviani |
  |:-----------------:|:-----------------:|
  | `wbonat@ufpr.br`  | `walmes@ufpr.br`  |
date: >
  63^a^ RBras & 17^o^ SEAGRO</br>
  July 24--28, 2017</br>
  UFLA, Lavras/MG
output:
  slidy_presentation:
    css: config/style.css
    footer: McGLMs for experimental data, Bonat & Zeviani (2017)
    highlight: haddock
    includes:
      in_header: config/MathJax.html
---

# Motivational data sets

## $\texttt{iris}$ data set

  * $\texttt{iris}$ data set was collected by Anderson (1935) on three
    species of irises in the Gaspé Peninsula of Quebec, Canada.
  * This data set was employed by R. A. Fisher (1936)
    to introduce the method of discriminant analysis.
  * Easily available in $\texttt{R}$ through $\texttt{data(iris)}$.
  * Four response variables representing measurements (in cm) of parts
    of the flowers (length and width of sepal and petal).
  * One experimental factor with three levels ($\texttt{setosa}$,
    $\texttt{versicolor}$ and $\texttt{virginica}$) representing the
    species of iris.
  * How the species are related with flowers size?
  * Size is not a really precise response variable and can be understood
    as a function of the four measured response variables.

```{r, include = FALSE, results = "hide", message = FALSE}
source("./config/_setup.R")
library(lattice)
library(latticeExtra)
library(gridExtra)
library(ellipse)
library(corrplot)
library(car)
# library(mcglm)
```

```{r, include = TRUE, results = "hide", eval = FALSE}
data(iris)

scatterplotMatrix(~Sepal.Length + Sepal.Width +
                      Petal.Length + Petal.Width | Species,
                  data = iris,
                  smooth = FALSE,
                  reg.line = FALSE,
                  ellipse = TRUE,
                  by.groups = TRUE,
                  diagonal = "none",
                  legend.pos = "bottomleft")
```

```{r, include = TRUE, results = "hide"}
panel.ell <- function(x, y, ...) {
    panel.xyplot(x, y, ...)
    panel.ellipse(x, y, level = 0.5,
                  center.pch = 19, center.cex = 1, ...)
    panel.ellipse(x, y, level = 0.9,
                  center.pch = 19, center.cex = 1, ...)
}

splom(~iris[1:4],
      data = iris,
      auto.key = list(columns = 3,
                      title = "Species",
                      cex.title = 1.1),
      groups = Species,
      lower.panel = panel.ell,
      upper.panel = panel.ell)
```

## $\texttt{soya}$ data set

  * Experiment carried out in a vegetation house with soybeans.
  * Two plants by plot with three levels of the factor amount of
    water in the soil ($\texttt{water}$) and five levels of potassium
    fertilization ($\texttt{pot}$).
  * The plots were arranged in five blocks ($\texttt{block}$).
  * Three response variables were measured, namely, grain yield,
    number of seeds and number of viable peas per plant.
  * The main goal is to assess the effect of the experimental factors
    on the soya $\textbf{yield}$.
  * Soya $\textbf{yield}$ is only indirectly measured through the three measured
    response variables.
  * Response variables are of mixed types, i.e. grain yield is a
    continuous outcome while number of seeds and number of viable peas
    per plant are examples of count and binomial response variables.
  * Data set is available in $\texttt{R}$ through the $\texttt{mcglm}$
    package.

```{r, include = TRUE, results = "hide", message = FALSE}
# data(soya)
# soya$viablepeasP <- soya$viablepeas/soya$totalpeas

# splom(~soya[, c(1,4:5,8)],
#       groups = soya$water,
#       layout= c(NA, 1),
#       type = c("p"),
#       auto.key = TRUE)

data(soybeanwp, package = "wzRfun")
str(soybeanwp)

soybeanwp$viab <- with(soybeanwp, nvp/(nvp + nip))


splom(#~soybeanwp[, c(1, 4, 7, 10)],
      ~soybeanwp[, c(1, 4:7, 10)],
      groups = soybeanwp$water,
      # as.matrix = TRUE,
      layout = c(NA, 1),
      upper.panel = function(x, y, groups, ...) {
          panel.xyplot(x, y, groups = groups, ...)
          i <- 0
          by(cbind(x, y),
             INDICES = groups,
             FUN = function(m) {
                 i <<- i + 1
                 panel.smoother(x = m[, 1],
                                y = m[, 2],
                                span = 0.9,
                                alpha.se = 0.15,
                                col = trellis.par.get()$superpose.line$col[i])
             })
      },
      auto.key = list(columns = 3,
                      title = "Water content (%)",
                      cex.title = 1.1))

# str(soybeanwp)

# Fitting the same model to all response variables.
m0 <- lm(cbind(yield, w100, Kconc, tg, viab) ~
             block + water * factor(potassium),
         data = soybeanwp)

r <- residuals(m0)
```

```{r}
# TODO fazer essa coisa acontecer!
# panel.qqmath.splom <- function(x, ...) {
#     # panel.qqmath(x, ...)
#     yr <- current.panel.limits()$ylim
#     qq <- qqnorm(x, plot.it = FALSE)
#     # u <- ((x - min(x))/diff(x))
#     # panel.ecdfplot(x, ...)
#     panel.xyplot(qq$x, qq$y, ...)
#     # panel.rug(x = x)
# }
#
# splom(~r,
#       lower.panel = panel.ell,
#       upper.panel = panel.ell,
#       diag.panel = panel.qqmath.splom)

scatterplotMatrix(r,
                  smooth = FALSE,
                  reg.line = FALSE,
                  ellipse = TRUE,
                  by.groups = TRUE,
                  diagonal = "qqplot",
                  legend.pos = "bottomleft")
```
