---
title: "Multivariate analysis of non-Gaussian data (McGLM)"
subtitle: Part II
date: |
  62^a^ RBras & 17^o^ SEAGRO\
  July 24--28, 2017\
  UFLA, Lavras/MG
author: |
  ```{r, echo = FALSE, results = "asis"}
  source("config/yaml-authors.R", print.eval = TRUE)
  ```
lang: pt-BR
graphics: yes
header-includes:
  - \usepackage{palatino}
  - \usepackage{inconsolata}
  - \usepackage{color}
#-- Uncomment for beamer.
#   - \let\oldframe\frame
#   - \renewcommand\frame[1][allowframebreaks]{\oldframe[#1]}
#   - \setbeamertemplate{frametitle continuation}{}
# classoption:
#   - serif
#   - mathserif
#   - professionalfont
#   - aspectratio=169
bibliography: config/ref.bib
#csl: config/ABNT_UFPR_2011-Mendeley.csl
cls: config/elsevier-harvard.csl
---

## A more general setup

  * MANOVA can deal only with Gaussian data.
  * Homogeneity of variances and covariances.
  * Independent observations within response variables.
  * What do we need?
    - Generalized linear models (GLM) to deal with non-Gaussian data.
    - Flexibility to deal with different types of dependence structures.
    - Multivariate response variables, possibly of mixed types.
    - Ideal approach -- easy to specify, fit, interpret and teach.

```{r, include = FALSE}
source("config/_setup.R")
```
```{r, include = FALSE}
data(iris)
head(iris)
```

# Multivariate analysis of non-Gaussian data

## Outline

  * Goal 1: Make GLM more flexible to deal with:
    - Non-negative highly right-skewed data and continuous data with
      probability mass at zero [@Bonat:2017].
    - Under, equi, overdispersed, zero-inflated and heavy-tailed
      count data [@Bonat:2017a].
    - Bounded data [@Bonat:2017b].
  * Goal 2: Extend GLM to cGLM to deal with:
    - Mixed models [@Bates:2015].
    - Repeated measures [@Diggle:2002].
    - Time series [@Box:1994].
    - Spatial and space-time data [@Cressie:1999; @Cressie:2011].
    - Genetic and Twin data [@Sorensen:2007].
  * Goal 3: Extend cGLM to McGLM to deal with:
    - Multivariate response variables [@Bonat:2016].
    - Response variables of mixed types [@Bonat:2017c].
    - Identify the general linear model and MANOVA as special cases.
  * Goal 4: Extend multivariate hypotheses tests to non-Gaussian data.
  * Goal 5: Provide software implementation in $\textsf{R}$.

## Flexible Generalized Linear Models

  * Let $\boldsymbol{Y}$ be an $N \times 1$ response vector.
  * Let $\boldsymbol{X}$ be an $N \times k$ design matrix.
  * Let $\boldsymbol{\beta}$ be a $k \times 1$ parameter vector.
  * Orthodox GLM
    $$
    \begin{eqnarray}
      \label{modelGLM}
      \mathrm{E}(\boldsymbol{Y}) &=&
        \boldsymbol{\mu} = g^{-1}(\boldsymbol{X} \boldsymbol{\beta})
        \nonumber \\
      \mathrm{Var}(\boldsymbol{Y}) &=&
        \boldsymbol{\Sigma} =
          \mathrm{V}(\boldsymbol{\mu};p)^{\frac{1}{2}}
          (\tau_0 \boldsymbol{I})
          \mathrm{V}(\boldsymbol{\mu};p)^{\frac{1}{2}}.
          \quad \quad \quad    (1)
      \end{eqnarray}
    $$
  * $g$ is the link function.
  * $\mathrm{V}(\boldsymbol{\mu};p) =
    \mathrm{diag}(\vartheta(\boldsymbol{\mu};p))$ where
    $\vartheta(\boldsymbol{\mu};p)$ is the variance function.
  * $p$ and $\tau_0$ are the power and dispersion parameters,
    respectively.
  * $\boldsymbol{I}$ is an identity matrix.
  * The standard linear model is obtained for identity link and constant
    variance functions.

## Variance/dispersion functions

  * Tweedie family (variance function), characterized by
    $$
      \vartheta(\boldsymbol{\mu};p) = \mu^p.
    $$
  * Gaussian ($p = 0$), gamma ($p = 2$) and inverse Gaussian ($p = 3$).
  * Deals with symmetric, skewed and zero-inflated continuous outcomes.
  * Poisson-Tweedie family (dispersion function), characterized by
    $$
      \vartheta(\boldsymbol{\mu};p) = \mu + \mu^p.
    $$
  * Hermite ($p=0$), Neyman Type A ($p=1$), negative binomial ($p=2$).
  * Special form in (1)
    $$
      \boldsymbol{\Sigma} = \mathrm{diag}(\boldsymbol{\mu}) +
        \mathrm{V}(\boldsymbol{\mu};p)^{\frac{1}{2}}
        (\tau_0 \boldsymbol{I})
        \mathrm{V}(\boldsymbol{\mu};p)^{\frac{1}{2}}.
    $$
  * Deals with over-/under-dispersion and zero-inflated count outcomes.
  * Binomial family, characterized by
    $$
      \vartheta(\boldsymbol{\mu}) = \mu(1-\mu).
    $$
  * Deals with binary, binomial and proportion outcomes.
  * Extended binomial variance function,
    $$
      \vartheta(\boldsymbol{\mu}; p, q) = \mu^p(1-\mu)^q.
    $$
  * Extra flexibility to deal with binary, binomial and proportion
    outcomes.
  * Three sets of variance functions.
  * Deals with the most common outcomes types.
  * Power parameter identifies the distribution.
  * Estimation of power parameter (model selection).
  * Additional flexibility.
  * Deals only with independent observations.

## Covariance Generalized Linear Models

  * Change the identity matrix in (1) to a non-diagonal matrix
    $\boldsymbol{\Omega}(\boldsymbol{\tau})$.
    $$
      \mathrm{Var}(\boldsymbol{Y}) =
        \boldsymbol{\Sigma} =
        \mathrm{V}(\boldsymbol{\mu};p)^{\frac{1}{2}}
          (\boldsymbol{\Omega}(\boldsymbol{\tau}))
          \mathrm{V}(\boldsymbol{\mu};p)^{\frac{1}{2}}.
    $$
  * Similar to working correlation matrix (GEE) [@Zeger:1988].
  * Model $\boldsymbol{\Omega}(\boldsymbol{\tau})$ as a linear
    combination of known matrices.
    $$
      h(\boldsymbol{\Omega}(\boldsymbol{\tau})) =
        \tau_0 Z_0 + \cdots + \tau_D Z_D,
    $$
    where $h$ is a covariance link function [@Pourahmadi:2000].
  * $Z_d$ with $d = 0, \ldots, D$ are known matrices.
  * $\boldsymbol{\tau} = (\tau_0,\cdots, \tau_D)$ is a $D \times 1$
    dispersion parameter vector.
  * Identity, inverse, exponential-matrix, modified Cholesky
    decomposition, etc.
  * Deals only with one response variable.
  * Examples of covariance linear models:
    - Double generalized linear models.
    - Linear mixed models.
    - Moving average models.
    - Exchangeable or compound symmetry and unstructured models (popular
      in longitudinal data analysis).
    - Conditional autoregressive models (time series, spatial and
      space-time data).
    - Models in quantitative genetic and phylogenetic.
    - Models for Twin and family data.
    - and many more.

## Example 1: Gaussian linear mixed model

  * Gaussian linear mixed model
    $$
      \boldsymbol{Y} =
        \boldsymbol{X} \boldsymbol{\beta} +
        \boldsymbol{U} \boldsymbol{\gamma} + \boldsymbol{\epsilon}
    $$
    where $\boldsymbol{\gamma}
    \sim N(\boldsymbol{0},\boldsymbol{D})$
    and $\boldsymbol{\epsilon}
    \sim N(\boldsymbol{0},\tau_0 \boldsymbol{I})$.
  * It is easy to show that
    $$
      \begin{eqnarray}
        \mathrm{E}(\boldsymbol{Y}) &=&
          \boldsymbol{\mu} = \boldsymbol{X} \boldsymbol{\beta}
          \nonumber \\
        \mathrm{Var}(\boldsymbol{Y}) &=&
          \boldsymbol{\Omega} =
          \tau_0 \boldsymbol{I} +
          \boldsymbol{U}\boldsymbol{D}\boldsymbol{U}^{\top}.
          \nonumber
      \end{eqnarray}
    $$
  * Furthermore,
    $$
      \boldsymbol{U}\boldsymbol{D}\boldsymbol{U}^{\top} =
        \sum_{j \leq q}^k d_{jq} \boldsymbol{V}_{jq}
    $$
    where
    $$
      \boldsymbol{V}_{jj} =
      \boldsymbol{U_{\cdot j}} \boldsymbol{U_{\cdot j}}^{\top}, \quad
      \boldsymbol{V}_{jq} = \boldsymbol{U_{\cdot j}}
        \boldsymbol{U_{\cdot q}}^{\top} +
        \boldsymbol{U_{\cdot q}}
        \boldsymbol{U_{\cdot j}}^{\top}
    $$
    for $j \neq q$ and $\boldsymbol{U_{\cdot j}}$ denotes the j$th$
    column of $\boldsymbol{U}$.

## Example 2: exchangeable and unstructured models

  * Exchangeable
    $$
      \begin{equation*}
        Z_1 = \begin{bmatrix}
                1 & 1 & 1\\
                1 & 1 & 1\\
                1 & 1 & 1
              \end{bmatrix}.
      \end{equation*}
    $$
  * Unstructured
    $$
      \begin{equation*}
        Z_1 = \begin{bmatrix}
                0 & 1 & 0\\
                1 & 0 & 0\\
                0 & 0 & 0
              \end{bmatrix}, \quad
        Z_2 = \begin{bmatrix}
                0 & 0 & 1\\
                0 & 0 & 0\\
                1 & 0 & 0
              \end{bmatrix} \quad \text{and} \quad
        Z_3 = \begin{bmatrix}
                0 & 0 & 0\\
                0 & 0 & 1\\
                0 & 1 & 0
              \end{bmatrix}.
        \end{equation*}
    $$

## Example 3: conditional autoregressive models

  * Specify the inverse of the covariance matrix
    $$
      \begin{equation*}
      \label{CAR}
        \boldsymbol{\Omega}^{-1}(\tau,\rho) =
          \tau (\boldsymbol{D} - \rho \boldsymbol{W}),
      \end{equation*}
    $$
    where $\boldsymbol{W}$ is a neighbourhood matrix and
    $\boldsymbol{D}$ is a diagonal matrix with the number of neighbours.
  * Linear covariance models using the inverse covariance link function
    $$
      \begin{equation*}
        \boldsymbol{\Omega}^{-1}(\boldsymbol{\tau}) =
        \tau_0 \boldsymbol{D} + \tau_1 \boldsymbol{W},
      \end{equation*}
    $$
    where $\tau_0 = \tau$ and $\tau_1 = - \tau \rho$.
  * $\boldsymbol{W}$ can describe temporal, spatial and spatial temporal
    neighborhoods.

## Example 4: genetic additive models

  - Let $\boldsymbol{A}$ denote an additive genetic relatedness matrix,
    $$
      \boldsymbol{\Omega}(\boldsymbol{\tau}) =
        \tau_0 \boldsymbol{I} + \tau_1 \boldsymbol{A},
    $$
    where $\tau_0$ and $\tau_1$ denote the environment and genetic
    variance components, respectively. For details see @Bonat:2017c.

## Multivariate Covariance Generalized Linear Models

  * Let $\boldsymbol{Y}_{N \times R} = \{\boldsymbol{Y}_1, \ldots,
    \boldsymbol{Y}_R\}$ be an outcome matrix.
  * Let $\boldsymbol{M}_{N \times R} = \{\boldsymbol{\mu}_1, \ldots,
    \boldsymbol{\mu}_R\}$ be an expected value matrix.
  * Let $\boldsymbol{\Sigma}_r =
      \mathrm{V}_r(\boldsymbol{\mu};p)^{\frac{1}{2}}
      \boldsymbol{\Omega}_r(\boldsymbol{\tau})
      \mathrm{V}_r(\boldsymbol{\mu};p)^{\frac{1}{2}}$ be the covariance
    matrix within outcomes.
  * Let $\boldsymbol{\Sigma}_b$ be the correlation matrix between
    outcomes.
  * We define the McGLM by,
    $$
      \begin{eqnarray}
        \mathrm{E}(\boldsymbol{Y}) &=&
          \boldsymbol{M} =
            \{g_1^{-1}(\boldsymbol{X}_1 \boldsymbol{\beta}_1),
            \ldots,
            g_R^{-1}(\boldsymbol{X}_R \boldsymbol{\beta}_R)\}
            \nonumber    \\
        \mathrm{Var}(\boldsymbol{Y}) &=&
          \boldsymbol{C} =
            \boldsymbol{\Sigma}_R \overset{G} \otimes
            \boldsymbol{\Sigma}_b
            \nonumber
      \end{eqnarray}
    $$
    where
    $$
      \boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b =
        \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots,
          \tilde{\boldsymbol{\Sigma}}_R)
        (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I})
        \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots,
        \tilde{\boldsymbol{\Sigma}}_R^\top).
    $$
  * Generalized Kronecker product proposed by @Martinez:2013.
  * $\tilde{\boldsymbol{\Sigma}}_r$ is the lower triangular matrix of
    the Cholesky decomposition of $\boldsymbol{\Sigma}_r$.
  * $\mathrm{Bdiag}$ denotes a block diagonal matrix.
  * General linear model is easily obtained by taking
    - Identity link functions.
    - $\boldsymbol{X}_r = \boldsymbol{X}$ for all $r$.
    - Constant variance function.
    - $\boldsymbol{\Sigma}_r = \tau_{0r} \boldsymbol{I}$ for all $r$.

# Estimation and Inference

## Model specification

  * Let $\mathcal{Y} = (\boldsymbol{Y}_1^\top, \ldots,
    \boldsymbol{Y}_R^\top)^\top$ be the stacked vector $(NR \times 1)$
    of the outcome matrix $\boldsymbol{Y}_{N \times R}$ by columns.
  * Let $\mathcal{M} = (\boldsymbol{\mu}_1^\top, \ldots,
    \boldsymbol{\mu}_R^\top)^\top$ be the stacked vector $(NR \times 1)$
    of the expected value matrix $\boldsymbol{M}_{N \times R}$ by
    columns.
  * The $\boldsymbol{C}$ matrix is symmetric and $NR \times NR$.
  * Let $\boldsymbol{\beta} = (\boldsymbol{\beta}_1^\top, \ldots,
    \boldsymbol{\beta}_R^\top)^\top$ be a vector $K \times 1$ of
    regression parameters.
  * Let $\boldsymbol{\lambda} = (\rho_1, \ldots, \rho_{R(R-1)/2}, p_1,
    \ldots, p_R, \boldsymbol{\tau}_1^\top, \ldots,
    \boldsymbol{\tau}_R^\top)^\top$ be a $Q \times 1$ vector of
    dispersion parameters.
  * McGLM are specified by two sets of parameters $\boldsymbol{\theta} =
    (\boldsymbol{\beta}, \boldsymbol{\lambda})$.
  * Quasi-score function for regression parameters.
  * Pearson estimating function for dispersion parameters.

## Estimating functions

  * The quasi-score function is defined by
    $$
      \begin{equation*}
        \psi_{\boldsymbol{\beta}}(\boldsymbol{\beta},
          \boldsymbol{\lambda}) = \boldsymbol{D}^\top
            \boldsymbol{C}^{-1}(\mathcal{Y} - \mathcal{M}),
      \end{equation*}
    $$
    where $\boldsymbol{D} = \nabla_{\boldsymbol{\beta}} \mathcal{M}$ is
    an $NR \times K$ matrix.
  * The Pearson estimating function is defined by,
    $$
      \begin{equation*}
        \label{Pearson}
        \psi_{\boldsymbol{\lambda}_i}(\boldsymbol{\beta},
        \boldsymbol{\lambda}) =
        \mathrm{tr}(W_{\boldsymbol{\lambda}i}
          (\boldsymbol{r}^\top\boldsymbol{r} -
          \boldsymbol{C})),
      \end{equation*}
    $$
    where $W_{\boldsymbol{\lambda}i} = -\frac{\partial
    \boldsymbol{C}^{-1}}{\partial \boldsymbol{\lambda}_i}$ and
    $\boldsymbol{r} = (\mathcal{Y} - \mathcal{M})$.

## Fitting algorithm

  * Modified chaser
    $$
    \begin{eqnarray}
    \label{chaser}
      \boldsymbol{\beta}^{(i+1)} &=&
        \boldsymbol{\beta}^{(i)} - \mathrm{S}_{\boldsymbol{\beta}}^{-1}
        \psi_{\boldsymbol{\beta}}
        (\boldsymbol{\beta}^{(i)}, \boldsymbol{\lambda}^{(i)})
        \nonumber \\
      \boldsymbol{\lambda}^{(i+1)} &=&
        \boldsymbol{\lambda}^{(i)} - \alpha
        \mathrm{S}_{\boldsymbol{\lambda}}^{-1}
        \psi_{\boldsymbol{\lambda}}(\boldsymbol{\beta}^{(i+1)},
          \boldsymbol{\lambda}^{(i)}). \nonumber
    \end{eqnarray}
    $$
  * $\mathrm{S}_{\boldsymbol{\beta}}$ and
    $\mathrm{S}_{\boldsymbol{\lambda}}$ are the sensitivity matrices
    for $\boldsymbol{\beta}$ and $\boldsymbol{\lambda}$,
    respectively.

## Godambe information matrix and asymptotic distribution

  * Denote $\hat{\boldsymbol{\theta}} = (\hat{\boldsymbol{\beta}},
    \hat{\boldsymbol{\lambda}})$ the estimating function estimator of
    $\boldsymbol{\theta}$.
  * The asymptotic distribution of $\hat{\boldsymbol{\theta}}$ is given
    by
    $$
    \begin{equation*}
      \hat{\boldsymbol{\theta}} \sim
        \mathrm{N}_{K+Q}(\boldsymbol{\theta},
          \mathrm{J}_{\boldsymbol{\theta}}^{-1})
    \end{equation*}
    $$
    where $\mathrm{J}_{\boldsymbol{\theta}}^{-1}$ is the inverse of the
    Godambe information matrix,
    $$
    \begin{equation*}
      \mathrm{J}_{\boldsymbol{\theta}}^{-1} =
        \mathrm{S}_{\boldsymbol{\theta}}^{-1}
        \mathrm{V}_{\boldsymbol{\theta}}
        \mathrm{S}_{\boldsymbol{\theta}}^{-\top},
    \end{equation*}
    $$
    where $\mathrm{S}_{\boldsymbol{\theta}}^{-\top} =
    (\mathrm{S}_{\boldsymbol{\theta}}^{-1})^{\top}$.
  * $\mathrm{V}_{\boldsymbol{\theta}}$ is the variability matrix.
  * For details, see @Bonat:2016 and @Jorgensen:2004.

## Multivariate linear hypotheses tests

  * Hypotheses
    $$
      H_{0}: \boldsymbol{L}\boldsymbol{\beta} = \boldsymbol{c}\quad
      \text{vs} \quad
      H_{1}:\boldsymbol{L}\boldsymbol{\beta} \neq \boldsymbol{c}.
    $$
  * Wald statistics
    $$
      W_{s} = (\boldsymbol{L}\boldsymbol{\beta})^{\top}
        (\boldsymbol{L}
          \mathrm{J}_{\boldsymbol{\beta}}^{-1}
          \boldsymbol{L})^{-1}
        (\boldsymbol{L}\boldsymbol{\beta}),
    $$
    where $\mathrm{J}_{\boldsymbol{\beta}} = \boldsymbol{D}^{\top}
    \boldsymbol{C}^{-1}\boldsymbol{D}.$
  * $\boldsymbol{L}$ is a $JR \times K$ known matrix of constant.
  * $J$ denotes the number of regression coefficients under testing.
  * Asymptotically $W_s$ has a $\chi^2$ distribution with $JR$ degrees
    of freedom.
  * $W_{s}$ is a straightforward generalization of the Hotelling-Lawley
    statistics to multivariate non-Gaussian data.

# Computational implementation in $\textsf{R}$

## Overview of the `mcglm` package.

  * Package `mcglm` available at `github` and CRAN.
  * Instalation
```{r, eval = FALSE}
library(devtools)

# From github
install_github("wbonat/mcglm")

# From CRAN
install.packages("mcglm")
```
  * Main function `mcglm`.
```{r}
library(mcglm)
args(mcglm)
```
  * Link functions: `logit`, `probit`, `cauchit`, `cloglog`, `loglog`,
    `identity`, `log`, `sqrt` and `inverse`.
  * Variance functions: `constant`, `tweedie`, `poisson_tweedie`,
    `binomialP` and `binomialPQ`.
  * Covariance link functions: `identity`, `inverse` and
    `exponential-matrix`.

### Linear covariance structures

| Functions    	| Description                                    	|
|:--------------|:--------------------------------------------------|
| `mc_id()`    	| Identity matrix                                	|
| `mc_ns()`    	| Unstructured model                             	|
| `mc_dglm()`  	| Double generalized linear models.              	|
| `mc_mixed()` 	| Linear mixed models (formula similar to lme4). 	|
| `mc_ma()`    	| Moving average models of order p.              	|
| `mc_rw()`    	| CAR models for times series.                   	|
| `mc_car()`   	| CAR models for space data.                     	|
| `mc_dist()`  	| Distance based models.                         	|
| `mc_twin()`  	| ACE, ADE, AE, and CE models for twin data.     	|

  * The users can use any list of symmetric matrices.
  * Combine pre-specified structures with new ones.

### Methods for `mcglm` objects

| Functions       | Description                                                |
|:----------------|:-----------------------------------------------------------|
| `print()`       | Simple printed display of model features.                  |
| `summary()`     | Standard regression output.                                |
| `fitted()`      | Fitted values for observed data.                           |
| `residuals()`   | Pearson, raw and standardized residuals.                   |
| `coef()`        | Coefficient estimates.                                     |
| `vcov()`        | Variance-covariance matrix of coefficient estimates.       |
| `confint()`     | Confidence intervals.                                      |
| `anova()`       | Analysis of variance tables for fitted models.             |
| `manova()`      | MANOVA-like test.                                          |
| `plot()`        | Diagnostic plots of Pearson residuals and algorithm check. |

### Extra features for `mcglm` objects

| Functions               | Description                        |
|:------------------------|:-----------------------------------|
| `gof()`                 | Measures of goodness-of-fit.       |
| `mc_sic()`              | SIC for regression parameters.     |
| `mc_sic_covariance()`   | SIC for dispersion parameters.     |
| `mc_bias_correct_std()` | Bias-corrected std.                |
| `mc_robust_std()`       | Robust std.                        |
| `mc_conditional_test()` | Conditional hypotheses tests.      |
| `mc_compute_rho()`      | Compute autocorrelation estimates. |
| `mc_initial_values()`   | Initial values for `mcglm`.        |

  - GOF's implemented: `plogLik`, `pAIC`, `pKLIC`, `pBIC`, `ESS`,
    `GOSHO` and `RJC`.

# Study cases

## Example 1: `iris` data set

```{r, message = FALSE}
# Loading extra packages and functions
library(mvtnorm)
library(Matrix)
library(mcglm)
library(wzRfun)
library(corrplot)
source("../review/functions.R")
```

```{r, message = FALSE}
#-----------------------------------------------------------------------
# MLM with lm() and manova().

data(iris)

# Standard general linear model.
fit1 <- lm(cbind(Sepal.Length,
                 Sepal.Width,
                 Petal.Length,
                 Petal.Width) ~ Species,
           data = iris)

#-----------------------------------------------------------------------
# General linear model in the `mcglm` package.

# Linear predictors.
linear_pred <- c(Sepal.Length ~ Species,
                 Sepal.Width  ~ Species,
                 Petal.Length ~ Species,
                 Petal.Width  ~ Species)

# Matrix linear predictor.
Z0 <- mc_id(iris)

# Fitting.
fit2 <- mcglm(linear_pred = linear_pred,
              matrix_pred = list(Z0, Z0, Z0, Z0),
              link = rep("identity", 4),
              variance = rep("constant", 4),
              data = iris,
              control_algorithm = list(correct = FALSE))

# Comparing lm() and mcglm().
logLik.mlm(fit1)
plogLik(fit2)

# Multivariate hypothesis test.
summary(manova(fit1),
        test = "Hotelling-Lawley")

# Multivariate hypothesis test using mcglm.
manova.mcglm(fit2)
```

## Example 2: `soybeanwp` data set

```{r, message = FALSE}
# rm(list = ls())

# Loading data set.
data(soybeanwp)

# Removes an outlier.
soybeanwp <- soybeanwp[-74,]

# Transform factors and create variables.
soybeanwp <- within(soybeanwp, {
    potassium <- as.factor(potassium)
    water <- as.factor(water)
    block <- as.factor(block)
    total <- nvp + nip
    viab <- nvp/total
})
str(soybeanwp)

# Setting the contrasts option to contr.sum.
options(contrasts = c(unordered = "contr.sum",
                      ordered = "contr.poly"))

#--------------------------------------------
# Standard MANOVA ignoring count and binomial data.

fit_man <- manova(cbind(yield, w100, Kconc, tg, viab) ~
                      block + water * potassium,
                  data = soybeanwp)

#--------------------------------------------
# McGLM.

# Linear predictors.
# Short code to create formulas.
# dput(names(soybeanwp))
form <- sprintf("%s ~ block + water * potassium",
                c("yield", "w100", "Kconc", "tg", "viab"))
form <- sapply(form, FUN = as.formula)
form

# Matrix linear predictor.
Z0 <- mc_id(soybeanwp)

# MANOVA fitted using mcglm.
r <- 5
fit_lm <- mcglm(linear_pred = c(form),
                matrix_pred = replicate(r, Z0, simplify = FALSE),
                link = rep("identity", each = r),
                variance = rep("constant", times = r),
                Ntrial = replicate(5, NULL),
                data = soybeanwp,
                control_algorithm = list(correct = FALSE))

# Fitting
fit_mcglm <- mcglm(linear_pred = c(form),
                   matrix_pred = list(Z0,
                                      Z0,
                                      Z0,
                                      Z0,
                                      Z0),
                   link = c("identity",
                            "identity",
                            "identity",
                            "log",
                            "logit"),
                   variance = c("constant",
                                "constant",
                                "constant",
                                "tweedie",
                                "binomialP"),
                   Ntrial = list(NULL,
                                 NULL,
                                 NULL,
                                 NULL,
                                 soybeanwp$total),
                   data = soybeanwp,
                   control_algorithm = list(correct = FALSE))

# Comparison using pseudo-logLik.
logLik.mlm(fit_man)
plogLik(fit_lm)
plogLik(fit_mcglm)

# MANOVA test.
# summary(fit_man, test = "Hotelling-Lawley")
car::Anova(fit_man, test = "Hotelling-Lawley", type = "III")
manova.mcglm(fit_lm)

# Multivariate hypothesis test.
manova.mcglm(fit_mcglm)

# Correlation among responses.
COR <- Matrix(0, 5, 5)
COR[lower.tri(COR)] <- fit_mcglm$Covariance[1:10]
COR <- matrix(forceSymmetric(COR, uplo = TRUE), 5, 5)
colnames(COR) <- rownames(COR) <- c("yield", "w100", "K", "tg", "viab")
```

```{r, fig.height = 7, fig.width = 7}
corrplot(COR,
         type = "upper",
         tl.pos = "d",
         outline = TRUE,
         method = "ellipse")
corrplot(COR,
         add = TRUE,
         type = "lower",
         method = "number",
         col = "black",
         diag = FALSE,
         tl.pos = "n",
         cl.pos = "n")
```

# Concluding remarks

## Discussion

  * Universal multivariate statistical modelling framework.
  * Flexible class of models based on second-moment assumptions.
  * General framework for estimation based on estimating function.
  * Efficient algorithms for estimation.
  * Asymptotic theory (Godambe Information).
  * Multivariate hypotheses tests for non-Gaussian data.

## Future work

### Extensions

  * Automatic covariate selection (Score Information Criterion-SIC).
  * Automatic covariance selection.
  * Measures of goodness-of-fit (Pseudo logLik, Pseudo AIC, Pseudo KLIC
    and ESS.)
  * Simulation from McGLMs based on the NORTA algorithm.

### Coming soon

  * Residual analysis (improvements).
  * Diagnostics (Leverage, DFBETA's and Cook's distance).
  * Penalized estimating functions (high dimensional data and splines).,
  * Prediction (time, space and space-time).
  * Special module for Twin data analysis. (NordicTwin cancer project)
  * Include parameter constraints to fit structural equation models
    (SEM).
  * Improve package documentation.
  * Multivariate hypotheses tests based on the pseudo-likelihood and
    score statistics.

# References

---

<!-- Coloca as referências aqui. -->
<div id="refs">
</div>

---

<center>
<p style='font-size: 80px;'>Thank you!</p>

<p>
  We are waiting you at Curitiba/PR for the 63<sup>a</sup> RBras!
  </br>
  Save the date: May 23 -- 25, 2018.
</p>
</center>

```{r, eval = FALSE, include = FALSE}
library(knitr)
# Include some postal places.
# include_graphics()
```
