---
title: "Computation and simulation study"
author: |
  ```{r, echo = FALSE, results = "asis"}
  source("../slides/config/yaml-authors.R", print.eval = TRUE)
  ```
date: >
  62^a^ RBras & 17^o^ SEAGRO</br>
  July 24--28, 2017</br>
  UFLA, Lavras/MG
bibliography: ../slides/config/ref.bib
csl: ../slides/config/elsevier-harvard.csl
---

# Fitting the MANOVA model using `lm()`

The data used below is from @pimentelgomes2002, tabel 10.2.1, page 254.
The measured responses are the content of nitrogen (`y1`) and phosphorus
(`y2`) in plants fertilized in pots. The experiment was installed in a
complete randomized design.

```{r, include = FALSE}
source("../slides/config/_setup.R")
# opts_chunk$set(eval = FALSE)
```
```{r}
#-----------------------------------------------------------------------
# Packages.

library(lattice)
library(latticeExtra)

#-----------------------------------------------------------------------
# Pimentel Gomes, page 254.

da <- expand.grid(rep = 1:5,
                  trt = gl(3, 1,
                           labels = c("Test", "TurFer", "TurNat")),
                  KEEP.OUT.ATTRS = TRUE)
da <- transform(da,
                y1 = c(463, 438, 494, 496, 448, 603, 596, 616, 633,
                       608, 471, 481, 449, 443, 456)/100,
                y2 = c(950, 890, 1010, 1230, 940, 1080, 1050, 1080,
                       1190, 1080, 960, 930, 870, 820, 910)/1000)
str(da)

# ATTENTION: This is a toy example, small dataset easy to handle.
xyplot(y2 ~ y1,
       groups = trt,
       auto.key = list(columns = 3,
                       title = "Treatments",
                       cex.title = 1),
       data = da,
       grid = TRUE) +
    glayer(panel.ellipse(...))

# Fitting the manova model (full model).
m_full <- lm(cbind(y1, y2) ~ trt, data = da)

# Summaries for separated univariate models.
summary(m_full)

# Separated ANOVA tables.
summary.aov(m_full)

# MANOVA table.
anova(m_full)

# Estimated parameters.
coef(m_full)

# Estimated covariance matrix (least squares).
var(residuals(m_full))

# Null model.
m_null <- update(m_full, . ~ 1)
```

# Calculating the statistics of the four MANOVA tests

```{r}
#-----------------------------------------------------------------------
# Doing the math.

# Error SSP of the full model.
E_full <- crossprod(residuals(m_full))
E_full

# Error SSP of the null model.
E_null <- crossprod(residuals(m_null))
E_null

# Extra error SSP = SSPnull - SSPfull.
E_extr <- E_null - E_full
E_extr

# Eigen values of the R^{-1} H.
lambdas <- eigen(solve(E_full, E_extr))$values

# Wilks' lambda.
det(E_full)/det(E_full + E_extr)
1/prod(1 + lambdas)
anova(m_full, test = "Wilks")

# Roy's largest root.
lambdas[1]
anova(m_full, test = "Roy")

# Hotelling-Lawley trace (Wald).
sum(diag(solve(E_full, E_extr)))
sum(lambdas)
anova(m_full, test = "Hotelling-Lawley")

# Pillai's trace.
sum(diag(E_extr %*% solve(E_extr + E_full)))
sum(lambdas/(1 + lambdas))
anova(m_full, test = "Pillai")
```

# A minimal of matrix algebra

```{r}
#-----------------------------------------------------------------------
# Calculating using matricial algebra.

# Matrix of responses.
Y <- as.matrix(da[, c("y1", "y2")])

# Total SSP = Y'Y.
t(Y) %*% Y

# Design matrix (full model).
X <- model.matrix(~trt, data = da)
dim(X)

# Estimation of the model parameters.
XlX <- crossprod(X)
XlY <- crossprod(X, Y)
B <- solve(XlX, XlY)
B

# Just to confirm.
coef(m_full)

# To create projection matrices.
proj <- function(X) {
    tX <- t(X)
    X %*% solve(tX %*% X) %*% tX
}

# Projection (hat) matrix for the full model.
Hfull <- proj(X)

# Projection (hat) matrix for the null model.
Hnull <- proj(X[, 1])

# Identity.
I <- diag(nrow(X))

# Error SSP of the full model.
E_full <- t(Y) %*% (I - Hfull) %*% Y
E_full

# Error SSP of the null model.
E_null <- t(Y) %*% (I - Hnull) %*% Y
E_null

# Extra error SSP.
E_extr <- t(Y) %*% (Hfull - Hnull) %*% Y
E_extr

# Eigen values of the R^{-1} H.
lambdas <- eigen(solve(E_full, E_extr))$values

# Wilks' lambda.
det(E_full)/det(E_full + E_extr)
1/prod(1 + lambdas)
anova(m_full, test = "Wilks")

# Roy's largest root.
lambdas[1]
anova(m_full, test = "Roy")

# Hotelling-Lawley trace (Wald).
sum(diag(solve(E_full, E_extr)))
sum(lambdas)
anova(m_full, test = "Hotelling-Lawley")

# Pillai's trace.
sum(diag(E_extr %*% solve(E_extr + E_full)))
sum(lambdas/(1 + lambdas))
anova(m_full, test = "Pillai")

# ATTENTION: models with more than one term, a QR decomposition of the X
# design matrix can be used.

#-----------------------------------------------------------------------
```

# The $\chi^2$ asymptotics

## Function prototyping

```{r}
#-----------------------------------------------------------------------
# Packages.

rm(list = ls())

library(mvtnorm)

#-----------------------------------------------------------------------
# Prototyping: creating a function to run the simulation study.

# Simulation settings.
m <- 3    # Number of responses.
k <- 4    # Number of groups.
rept <- 4 # Repetition of each group.

# Returns the chi-square statistic of each test.
simul_chi <- function(m, k, rept, N = 1000) {
    trt <- gl(k, rept)
    n <- length(trt)
    # Means and covariances.
    mu <- rep(0, m)
    Sigma <- 0.5 * (diag(m) + matrix(1, m, m))
    # Paired samples of the statistics.
    smp <- replicate(N, {
        Y <- rmvnorm(n = length(trt),
                     mean = mu,
                     sigma = Sigma)
        # Sigma of the null model.
        m0 <- lm(Y ~ 1)
        S0 <- var(residuals(m0)) * (n - 1)/n
        # Sigma of the full model.
        m1 <- lm(Y ~ trt)
        S1 <- var(residuals(m1)) * (n - 1)/n
        #------------------------------------
        # Wilks' statistic.
        logdet <- log(det(S1)/det(S0))
        # Has an asymptotic chi-square aproxiamation.
        wil <- -n * logdet
        # ATTENTION: this has a better chi-square approximation.
        # wil <- -(n - k - 1/2 * (m - (k - 1) + 1)) * logdet
        #------------------------------------
        # Hotelling-Lawley's statistic.
        hol <- n * sum(diag(solve(S1, S0 - S1)))
        #------------------------------------
        # Pillai's statistic.
        pil <- n * sum(diag(solve(S0, S0 - S1)))
        return(c(wil = wil, hol = hol, pil = pil))
    })
    return(smp)
}

# A sample of the Wilks' lambda statistic.
smp <- simul_chi(m = m, rept = rept, k = k)

# The function to display the legend.
leg <- function(...) {
    legend(...,
           legend = c("Wilks", "Hotelling-Lawley", "Pillai"),
           col = c(2, 4, 3),
           lty = 1,
           bty = "n")
}

# Empirical chi-square distribution vs theoretical.
plot(ecdf(smp[1, ]), col = 2,
     xlab = expression(X^{2} ~ "statistic"),
     ylab = "Empirical CDF",
     main = NULL)
lines(ecdf(smp[2, ]), col = 4)
lines(ecdf(smp[3, ]), col = 3)
curve(pchisq(x, df = m * (k - 1)), add = TRUE, col = 1)
leg("right")

# Get the p-values.
pval <- apply(smp,
              MARGIN = 1,
              FUN = pchisq,
              df = m * (k - 1))
str(pval)

plot(ecdf(pval[, 1]),
     xlab = expression("p-values for the" ~ X^{2} ~ "statistic"),
     ylab = "Empirical CDF",
     col = 2,
     pch = NA,
     main = NULL)
lines(ecdf(pval[, 2]), col = 4, cex = NA)
lines(ecdf(pval[, 3]), col = 3, cex = NA)
segments(0, 0, 1, 1, col = 1)
leg("left")
```

## Running the simulation study

```{r}
#-----------------------------------------------------------------------
# Design settings for a small simulation study.

# k:    number of treatment levels (fixed);
# N:    number of simulations (fixed);
# m:    number of responses (varying);
# rept: repetition per treatment level (varying).
k <- 4
N <- 1000
ds <- expand.grid(m = c(2, 5, 8),
                  rept = c(5, 10, 20))

# Simulation for each design setting case.
res <- by(data = ds,
          INDICES = seq_len(nrow(ds)),
          FUN = function(case) {
              smp <- with(case,
                          simul_chi(m = m,
                                    rept = rept,
                                    k = k,
                                    N = N))
              test <- rownames(smp)
              smp <- c(t(smp))
              data.frame(m = case$m,
                         rept = case$rept,
                         test = rep(test, each = N),
                         smp = smp,
                         pval = pchisq(smp, case$m * (k - 1)))
          })

# Stack to one data frame.
res <- do.call(rbind, res)
res$n <- res$rept * k
res$test <- factor(res$test,
                   levels = c("wil", "hol", "pil"),
                   labels = c("Wilks", "Hotelling-Lawley", "Pillai"))
res_chi <- res
rm(res)

useOuterStrips(
    ecdfplot(~pval | factor(m) + factor(n),
             as.table = TRUE,
             groups = test,
             auto.key = list(columns = 2,
                             title = "Test",
                             cex.title = 1.1),
             xlab = expression("p-values for the" ~
                                   X^{2} ~ "statistic"),
             ylab = "Empirical CDF",
             data = res_chi),
    strip = strip.custom(
        var.name = "Responses",
        strip.names = TRUE,
        sep = " = "),
    strip.left = strip.custom(
        var.name = "Subjects",
        strip.names = TRUE,
        sep = " = ")) +
    layer({
        panel.segments(0, 0, 1, 1, lty = 2, col = 1)
    })
```

# The $F$ approximation

## Running the simulation study

```{r}
#-----------------------------------------------------------------------
# The F approximantion of each multivariate test.

# Each test.
tt <- c("Wilks", "Hotelling-Lawley", "Pillai", "Roy")
# tt <- tt[-4]

# Simulation of data under H_0 and computation of F statistics.
simul_F <- function(m, k, rept, N = 1000) {
    trt <- gl(k, rept)
    mu <- rep(0, m)
    n <- length(trt)
    Sigma <- 0.5 * (diag(m) + matrix(1, m, m))
    # Sample of the statistic.
    smp <- replicate(N, {
        Y <- rmvnorm(n = length(trt),
                     mean = mu,
                     sigma = Sigma)
        m0 <- lm(Y ~ trt)
        pvals <- sapply(tt,
                        FUN = function(test) {
                            # Extract the p-value.
                            anova(m0, test = test)[2, 6]
                        })
        return(pvals)
    })
    return(smp)
}

# Testing.
simul_F(m = 3, k = 4, rept = 10, N = 5)

# Simulation for each design setting case.
res <- by(data = ds,
          INDICES = seq_len(nrow(ds)),
          FUN = function(case) {
              smp <- with(case,
                          simul_F(m = m,
                                  rept = rept,
                                  k = k,
                                  N = N))
              smp <- stack(as.data.frame(t(smp)))
              cbind(data.frame(m = case$m,
                               rept = case$rept),
                    smp)
          })

# Stack to one data frame.
res <- do.call(rbind, res)
res$n <- res$rept * k
names(res)[3:4] <- c("pval", "test")
res$test <- factor(res$test, levels = tt)
res_F <- res
rm(res)

useOuterStrips(
    ecdfplot(~pval | factor(m) + factor(n),
             groups = test,
             auto.key = list(columns = 2,
                             title = "Test",
                             cex.title = 1.1),
             as.table = TRUE,
             xlab = expression("p-values for the" ~
                                   italic(F) ~ "statistic"),
             ylab = "Empirical CDF",
             subset = !is.na(pval),
             data = res_F),
    strip = strip.custom(
        var.name = "Responses",
        strip.names = TRUE,
        sep = " = "),
    strip.left = strip.custom(
        var.name = "Subjects",
        strip.names = TRUE,
        sep = " = ")) +
    layer({
        panel.segments(0, 0, 1, 1, lty = 2, col = 1)
    })

#-----------------------------------------------------------------------
```

# The Hotelling-Lawley trace and the Wald test

```{r, include = TRUE, eval = TRUE}
rm(list = ls())

# The iris dataset.
m0 <- lm(cbind(Petal.Length,
               Petal.Width,
               Sepal.Length,
               Sepal.Width) ~ Species,
         data = iris)

# The model matrix and the X'X.
X <- model.matrix(m0)
XlX <- crossprod(X)

B <- coef(m0)    # Matrix of parameters.
b <- cbind(c(B)) # vec(B): vectorized form.

n <- nrow(X)  # Number of sample units.
k <- nrow(B)  # Number of lines in B or columns in X.
r <- ncol(B)  # Number of responses.

# To test H_0: the Spiecies are equal in all responses.
A <- rbind(c(0, 1, 0),
           c(0, 0, 1))
h <- nrow(A)  # Number of linear functions.

# All responses.
M <- cbind(diag(r))

# Only the first response.
# M <- cbind(diag(r)[, 1])

v <- ncol(M) # Number of linear combination of responses.

# Breads, ham and the Sandwich (H = hypothesis SSP).
ABM <- A %*% B %*% M
AXlXiA <- A %*% solve(XlX) %*% t(A)
H <- t(ABM) %*% solve(AXlXiA) %*% ABM
H

library(car)

# The F-statistic for the ABM = 0 hypothesis.
linearHypothesis(m0,
                 hypothesis.matrix = A,
                 P = M)

# The residual SPP of the full model.
R <- crossprod(residuals(m0))

# Model under the constrain of the H_0.
m00 <- update(m0, . ~ 1)

# The hypothesis SPP as a difference of residuals SPP.
H <- crossprod(residuals(m00)) - R
H

# Function that returns the trace of a matrix.
tr <- function(x) sum(diag(x))
tr(solve(R) %*% H)
tr(H %*% solve(R))

# The trace is the Hotelling-Lawley criterion.
anova(m0, test = "Hotelling")

# Testing the Kronecker operation in R.
# kronecker(diag(1:4), matrix(1, 2, 2))

# Testing the same hypothesis but with the vectorized components.
L <- kronecker(t(M), A)
b
Lb <- L %*% b
Lb

# Sigma <- var(residuals(m0))
Sigma <- R/n

# MSM <- t(M) %*% Sigma %*% M
# MSM.XlXi <- kronecker(MSM, solve(XlX))
# (1/n) * t(Lb) %*% solve(L %*% MSM.XlXi %*% t(L)) %*% Lb

# Wald statistic.
t(Lb) %*%
    solve(L %*%
          kronecker(t(M) %*% R %*% M,
                    solve(XlX)) %*%
          t(L)) %*%
    Lb
# H-L trace.
tr(H %*% solve(R))

# So, Wald is the Hotelling-Lawley trace criterion.
```

# References

```{r, echo = FALSE, results = "asis", message = FALSE, warning = FALSE, error = FALSE}
if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "html") {
    cat('<!-- Insert the refereces right here! -->',
        '<div id="refs">',
        '</div>\n',
        sep = "\n")
}
```

# Session information

```{r}
# devtools::session_info()
Sys.time()
cbind(Sys.info())
sessionInfo()
```
