---
title: "Soil chemical variables in samples on a *Tectona grandis* plantation"
author: |
  ```{r, echo = FALSE, results = "asis"}
  source("../slides/config/yaml-authors.R", print.eval = TRUE)
  ```
date: >
  62^a^ RBras & 17^o^ SEAGRO</br>
  July 24--28, 2017</br>
  UFLA, Lavras/MG
bibliography: ../slides/config/ref.bib
csl: ../slides/config/elsevier-harvard.csl
---

```{r, include = FALSE}
source("../slides/config/_setup.R")
```

# Data description and objectives

The dataset contains the values of chemical variables of 3 layers of the
soil in 50 sites cultivated with teak (*Tectona grandis*). The study was
carried out in 2015, in Teca plantations belonging to two farms located
in the Western Region of the State of Mato Grosso. The selection of
study areas within the farms was done by means of free walking, covering
the entire cultivated area of 1869 ha, making field observations and
delimiting plots within the stands from the characteristics of the
soils, position in the landscape and development of culture. Fifty plots
with 600 m^2^ (20 x 30 m) each were allocated. As a selection criterion,
only fields with an area of more than seven hectares (7 ha) were
selected, using only the areas with the same planting density,
management practices aged 13-14 years.

The objective of the study is to conduct a profile analysis to verify if
there is a general pattern of cation concentration (K^+^, Ca^2+^ and
Mg^2+^) in the soil layers.

This dataset is part of data base with more tables (`teca_arv`,
`teca_crapar`, `teca_cra`, `teca_gran` and `teca_qui`). These data are
available on the package `EACS`
(http://leg.ufpr.br/~walmes/pacotes/EACS/index.html). Please do not use
datasets without the researchers' permission because the results of
these researches are being submitted to journals at the moment. If you
want to use the data to propose statistical methodologies, contact the
researchers as well.

# Analysis with the multivariate linear model (MLM)

## Loading the dataset

```{r, message = FALSE}
#-----------------------------------------------------------------------
# Packages.

library(latticeExtra)
library(car)
library(reshape2)
library(corrplot)
library(grid)
library(gridExtra)
```
```{r, eval = FALSE}
#-----------------------------------------------------------------------
# Getting the dataset.

# Online documentation of the EACS::teca_qui dataset.
u <- "http://leg.ufpr.br/~walmes/pacotes/EACS/reference/teca_qui.html"
browseURL(u)
```
```{r}
csv <- "https://raw.githubusercontent.com/walmes/EACS/master/data-raw/teca_qui.csv"
teca <- read.csv2(file = csv, dec = ".")
str(teca)

teca$cam <- factor(teca$cam, labels = 1:3)
```

## Exploratory data analysis

```{r}
#-----------------------------------------------------------------------
# Graphical exploratory analysis.

# The cations.
v <- c("k", "ca", "mg")
summary(teca[v])

# Only the cations.
densityplot(~k + ca + I(mg + 0.1),
            outer = TRUE,
            groups = cam,
            scales = list(relation = "free",
                          x = list(log = 10)),
            as.table = TRUE,
            data = teca)

scatterplotMatrix(~log(k) + log(ca) + log(mg + 0.1) | cam,
                  data = teca,
                  gap = 0,
                  smooth = FALSE,
                  reg.line = FALSE,
                  ellipse = TRUE,
                  by.groups = TRUE,
                  diagonal = "qqplot")

# Transformed copies of the variables.
teca <- transform(teca,
                  lk = log(k),
                  lca = log(ca),
                  lmg = log(mg + 0.1))

# Names of the transformed variables.
v <- tail(names(teca), n = 3)
v

#-----------------------------------------------------------------------
# Repeated measures design.

# Outer factor: none.
# Inner factor: soil layer (`cam`)
# Responses: 3 cations x 3 layers = 9 conditions.

# Long format.
tecal <- melt(data = teca[c("loc", "cam", v)],
              measure.vars = v,
              variable.name = "res")
str(tecal)

bwplot(value ~ cam | res,
       pch = "|",
       # scales = list(y = list(relation = "free")),
       layout = c(NA, 1),
       xlab = "Soil layer",
       ylab = "Observed values in the transformed scale",
       data = tecal) +
    layer(panel.xyplot(x = x,
                       y = y,
                       jitter.x = TRUE,
                       type = c("p", "a")))

# Combine 3 responses x 3 layers = 9 response variables.
tecal$res.cam <- with(tecal, paste(res, cam, sep = "."))

# Wide format.
tecaw <- dcast(data = tecal,
               formula = loc ~ res.cam,
               value = "value",
               stringsAsFactors = FALSE)

# Presented order of the responses.
ord <- c(t(outer(Y = levels(tecal$cam),
                 X = v,
                 FUN = paste,
                 sep = ".")))

tecaw <- tecaw[, c("loc", ord)]
str(tecaw)
```

## Repeated measures analysis

```{r}
#-----------------------------------------------------------------------
# Repeated measures analysis.

dput(names(tecaw)[-1])

# Multivariate linear model for 9 responses.
m0 <- lm(as.matrix(tecaw[, 2:10]) ~ 1)
m0

# summary(m0)
# summary.aov(m0)
anova(m0)

# Extract the raw residuals.
r <- residuals(m0)

# Checking the models assumptions on the residuals.
scatterplotMatrix(r,
                  gap = 0,
                  smooth = FALSE,
                  reg.line = FALSE,
                  ellipse = TRUE,
                  diagonal = "qqplot")

corrplot(cor(r),
         type = "upper",
         tl.pos = "d",
         outline = TRUE,
         method = "ellipse")

# Inner factors data design.
# DANGER this order changes de SS.
idata <- expand.grid(cam = levels(tecal$cam),
                     res = v,
                     KEEP.OUT.ATTRS = FALSE)
str(idata)

Anova(m0,
      idata = idata,
      idesign = ~res * cam)
```

## Testing linear hypotheses on the responses

```{r}
#-----------------------------------------------------------------------

# Regression parameters.
B <- coef(m0)
colnames(B)

# The L matrix is only a scalar.
L <- matrix(1, nrow = 1, ncol = 1)

# Check the orders.
cbind(idata, res.cam = cbind(colnames(B)))

# To test effect of responses.
M_res <- cbind(c(-1, -1, -1, -1, -1, -1, 2, 2, 2),
               c(-1, -1, -1,  1,  1,  1, 0, 0, 0))

linearHypothesis(m0,
                 test = "Pillai",
                 hypothesis.matrix = L,
                 P = M_res)

# To test effect of soil layer.
M_cam <- cbind(c(-1, -1, 2, -1, -1, 2, -1, -1, 2),
               c(-1,  1, 0, -1,  1, 0, -1,  1, 0))

linearHypothesis(m0,
                 test = "Pillai",
                 hypothesis.matrix = L,
                 P = M_cam)

# GOOD: To easy get the M matrices, creates a model matrix.
M <- model.matrix(~res * cam,
                  data = idata,
                  contrasts = list(res = contr.sum,
                                   cam = contr.sum))
a <- attr(M, "assign")

# This design matrix has all the slices needed.
M

# ATTENTION: The order or lines in the `idesign` and in the `B`
# parameter matrix must match.

linearHypothesis(m0,
                 test = "Pillai",
                 hypothesis.matrix = L,
                 P = M[, a == 3])
```

## An univariate specification

```{r}
#-----------------------------------------------------------------------
# The univariate regression model assumes independent observations
# (between responses and between soil layers) inside each location (a
# strong assumption in this case).

an0 <- lm(value ~ res * cam, data = tecal)
anova(an0)

# The point estimates for means are the same.
doBy::LSmeans(an0, effect = c("cam", "res"))
cbind(c(coef(m0)))
```

# Analysis with the McGLM

## Package instalation from GitHub

```{r, eval = FALSE}
# Installing the package from the github repository.
devtools::install_github("wbonat/mcglm", ref = "devel")
```
```{r}
# Load the package.
library(mcglm)
# packageVersion("mcglm")

# Sort the lines per subject (`loc`) followed by `cam`.
teca <- plyr::arrange(teca, loc, cam)

# Keep only the variables that will be used.
teca <- teca[, c(1, 2, 16, 17, 18)]
str(teca)

# Head and tail of the data.frame.
head(teca)
tail(teca)
```

## Covariance structures for the repeated measures

```{r}
#-----------------------------------------------------------------------
# Covariance structures for the inner subject factor `cam`.

# Independent covariance structure.
Z0 <- mc_id(teca)
length(Z0)

# Unstructured model for covariance among cam.
Z_ns <- mc_ns(teca, id = "loc")
length(Z_ns)

# Moving average first order structure.
Z_ma1 <- mc_ma(id = "loc",
               time = "cam",
               data = teca,
               order = 1)
length(Z_ma1)

# Distance based covariance structure.
Z_dist <- mc_dist(id = "loc",
                  time = "cam",
                  data = teca,
                  method = "euclidean")
length(Z_dist)

# Random walk covariance structure.
Z_rw <- mc_rw(id = "loc",
              time = "cam",
              data = teca,
              order = 1,
              proper = TRUE)
length(Z_rw)

#-----------------------------------------------------------------------
# The covariance structures.

# display.brewer.all()
colr <- brewer.pal(9, "GnBu")
colr <- colorRampPalette(colr, space = "rgb")

i <- 1:9
j <- i[i %% 3 == 0][-1] - 2.5
myplot <- function(m, ...) {
    levelplot(m, col.regions = colr, ...) +
        layer(panel.abline(v = j, h = j))
}
```

```{r, fig.height = 7 * 3/2}
grid.arrange(myplot(as.matrix(Z0[[1]])[i, i],
                    main = "Independent"),
             myplot(as.matrix(Z_ns[[1]] +
                              0.5 * Z_ns[[2]] +
                              0.75 * Z_ns[[3]])[i, i],
                    main = "Unstructured"),
             myplot(as.matrix(Z_ma1[[1]])[i, i],
                    main = "First order moving average"),
             myplot(as.matrix(Z_dist[[1]])[i, i],
                    main = "Euclidian distance"),
             myplot(solve(as.matrix(Z_rw[[1]] +
                                    0.5 * Z_rw[[2]])[i, i]),
                    main = "Random walk"),
             ncol = 2)
```

## Fitting several models

```{r}
#-----------------------------------------------------------------------
# Fitting several competing models.

# Linear predictors (the same RHS for all responses).
form <- list(lk ~ cam,
             lca ~ cam,
             lmg ~ cam)

# form00 <- c( lk.1 ~ 1,  lk.2 ~ 1,  lk.3 ~ 1,
#             lca.1 ~ 1, lca.2 ~ 1, lca.3 ~ 1,
#             lmg.1 ~ 1, lmg.2 ~ 1, lmg.3 ~ 1)

# To easy create a list of formulas.
form00 <- sapply(sprintf("%s ~ 1",
                         names(tecaw)[-1]),
                 FUN = as.formula)

# Independent subjects covariance structure.
Z00 <- mc_id(tecaw)

# MANOVA for all responses (wide format, subject is a single line).
fit00 <- mcglm(linear_pred = form00,
               matrix_pred = replicate(length(form00),
                                       Z00,
                                       simplify = FALSE),
               data = tecaw)

# Standard MANOVA (ignores the correlation among soil layers).
fit1 <- mcglm(linear_pred = form,
              matrix_pred = list(Z0, Z0, Z0),
              data = teca)

# MANOVA + repeated measures using unstructured matrix.
fit2 <- mcglm(linear_pred = form,
              matrix_pred = list(c(Z0, Z_ns),
                                 c(Z0, Z_ns),
                                 c(Z0, Z_ns)),
              control_algorithm = list(tunning = 0.9,
                                       max_iter = 100),
              data = teca)

# MANOVA + repeated measures using moving average first order.
fit3 <- mcglm(linear_pred = form,
              matrix_pred = list(c(Z0, Z_ma1),
                                 c(Z0, Z_ma1),
                                 c(Z0, Z_ma1)),
              control_algorithm = list(tunning = 0.8),
              data = teca)

# MANOVA + repeated measures using distance based.
fit4 <- mcglm(linear_pred = form,
              matrix_pred = list(c(Z0, Z_dist),
                                 c(Z0, Z_dist),
                                 c(Z0, Z_dist)),
              control_algorithm = list(tunning = 0.8),
              data = teca)

# MANOVA + repeated measures using distance based + expm covariance link
# function.
fit5 <- mcglm(linear_pred = form,
              matrix_pred = list(c(Z0, Z_dist),
                                 c(Z0, Z_dist),
                                 c(Z0, Z_dist)),
              covariance = c("expm",
                             "expm",
                             "expm"),
              control_algorithm = list(tunning = 0.8),
              data = teca)

# MANOVA + repeated measures using random walk + expm covariance link
# function.
fit6 <- mcglm(linear_pred = form,
              matrix_pred = list(c(Z_rw),
                                 c(Z_rw),
                                 c(Z_rw)),
              covariance = c("inverse",
                             "inverse",
                             "inverse"),
              control_algorithm = list(tunning = 0.5, max_iter = 100),
              data = teca)
```

## Comparing the models

```{r}
#-----------------------------------------------------------------------
# Comparing models fit.

m <- c("MANOVA",
       "Indep",
       "Unstr",
       "MA(1)",
       "Eucl",
       "Eucl expm",
       "RW expm")
meas <- rbind(gof(fit00),
              gof(fit1),
              gof(fit2),
              gof(fit3),
              gof(fit4),
              gof(fit5),
              gof(fit6))
meas <- cbind(model = m,
              as.data.frame(meas),
              stringsAsFactors = FALSE)
meas$model <- with(meas,
                   factor(model,
                          levels = model[order(Df)]))
meas <- plyr::arrange(meas, Df)
meas
```
```{r, fig.height = 12}
barchart(plogLik + pAIC + pBIC + pKLIC~ model,
         outer = TRUE,
         layout = c(1, NA),
         as.table = TRUE,
         ylab = "Measures of model fitting",
         xlab = "Fitted models",
         data = meas,
         scales = "free") +
    layer(grid.text(x = unit(1:nlevels(x), "native"),
                    y = unit(0.03, "npc"),
                    label = meas$Df))
```

## Pillai test for McGLM

```{r}
#-----------------------------------------------------------------------
# Multivariate hypotheses tests for McGLM.

source("../review/functions.R")
library(Matrix)

manova.mcglm(fit1)
manova.mcglm(fit2)
```

## Estimated correlations

```{r}
#--------------------------------------------
# Correlations.

# Correlation between responses.
summary(fit2, print = "Correlation")

# Correlation among `cam` for the `lk` response.
COR_lk <- matrix(NA, 3, 3)
COR_lk[lower.tri(COR_lk)] <- fit2$Covariance[5:7]
COR_lk[upper.tri(COR_lk)] <- fit2$Covariance[5:7]
diag(COR_lk) <- fit2$Covariance[4]
cov2cor(COR_lk)

# Correlation among `cam` for the `lca` response.
COR_lca <- matrix(NA, 3, 3)
COR_lca[lower.tri(COR_lca)] <- fit2$Covariance[9:11]
COR_lca[upper.tri(COR_lca)] <- fit2$Covariance[9:11]
diag(COR_lca) <- fit2$Covariance[8]
cov2cor(COR_lca)

# Correlation among `cam` for the `lmg` response.
COR_lmg <- matrix(NA, 3, 3)
COR_lmg[lower.tri(COR_lmg)] <- fit2$Covariance[13:15]
COR_lmg[upper.tri(COR_lmg)] <- fit2$Covariance[13:15]
diag(COR_lmg) <- fit2$Covariance[12]
cov2cor(COR_lmg)

# TODO How create the 9 x 9 correlogran using these values?
```

# References

```{r, echo = FALSE, results = "asis", message = FALSE, warning = FALSE, error = FALSE}
if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "html") {
    cat('<!-- Insert the refereces right here! -->',
        '<div id="refs">',
        '</div>\n',
        sep = "\n")
}
```

# Session information

```{r}
# devtools::session_info()
Sys.time()
cbind(Sys.info())
sessionInfo()
```
