---
title: "Water and potassium on the production of soybean"
author: |
  | Wagner H. Bonat   | Walmes M. Zeviani |
  |:-----------------:|:-----------------:|
  | `wbonat@ufpr.br`  | `walmes@ufpr.br`  |
  |  LEG/DEST/UFPR    |  LEG/DEST/UFPR    |
date: >
  62^a^ RBras & 17^o^ SEAGRO</br>
  July 24--28, 2017</br>
  UFLA, Lavras/MG
bibliography: ../slides/config/ref.bib
csl: ../slides/config/elsevier-harvard.csl
---

```{r, include = FALSE}
source("../slides/config/_setup.R")
```

# Data description and objectives

The tropical soils, usually poor in potassium (K), demand potassium
fertilization when cultivated with soybean (*Glycine max* L.) to obtain
satisfactory yields. The aim of this study was to evaluate the effects
of K doses (`pts`) and soil humidity (`wtr`) levels on soybean agronomic
characteristics.

The experiment was carried out in a greenhouse, in pots with two plants,
containing 5 dm^3^ of soil. The experimental design was completely
randomized block with treatments in a 5 $\times$ 3 factorial
arrangement. The K doses were 0; 30; 60, 120 and 180 mg dm^-3^ and the
soil humidity ranged from 35 to 40; 47.5 to 52.5; and 60 to 65% of the
total porosity. The responses measured were: grain yield
(`yield`), weight of a hundred grains (`w100`), total number of grains
per pot (`tg`), K level in the grain (`Kconc`), number of viable pods
(`nvp`, `nip` is the inviable pods).

For more information, see @Serafim2012.

```{r, message = FALSE}
#-----------------------------------------------------------------------
# Packages.

rm(list = ls())
library(lattice)
library(latticeExtra)
library(car)
library(candisc)
library(corrplot)
library(doBy)
library(multcomp)
```

# MLM for the soybean dataset

```{r}
#-----------------------------------------------------------------------
# Multivariate analysis of the soybean dataset.

# Visit to see the online documentation:
# https://github.com/walmes/wzRfun/blob/master/R/wzRfun.R#L10.

data(soybeanwp, package = "wzRfun")
str(soybeanwp)
head(soybeanwp)

# A copy with shorter names.
swp <- soybeanwp
# dput(names(swp))
names(swp) <- c("pts", "wtr", "blk",
                "yield", "w100", "Kconc", "tg", "nip", "nvp")

# Convert the controled numeric factors to categorical factors.
swp <- transform(swp,
                 Pts = factor(pts),
                 Wtr = factor(wtr))

# Creates the proportion of viable pods.
swp$pvp <- with(swp, nvp/(nvp + nip))
str(swp)

#-----------------------------------------------------------------------
# Data visulization.

# The 74 observation is an outlier to yield and tg.

combineLimits(
    useOuterStrips(
        xyplot(yield + tg + w100 + Kconc + pvp ~ pts | Wtr,
               outer = TRUE,
               as.table = TRUE,
               # data = swp[-74, ],
               data = swp,
               pch = 19,
               xlab = "Potassium",
               ylab = "Values",
               scales = list(y = "free"),
               groups = blk))) +
    layer(panel.xyplot(x = x, y = y, type = "a", col = 1))

# Data summaries.
summary(swp)
```

## Analysing the subset for potassium equals 0

```{r}
#-----------------------------------------------------------------------
# Condition 1: no sufficient evidence in each response.

# Subsetting the dataset.
da <- subset(swp[-74, ], pts == 0)

xtabs(~blk + Wtr, data = da)

xyplot(yield + tg + w100 + Kconc + pvp ~ wtr,
       outer = TRUE,
       scales = "free",
       type = c("p", "a"),
       data = da)

splom(~da[c(4:7, 12)],
      data = da,
      groups = Wtr,
      as.matrix = TRUE)

# `Wtr` is a 3 level categorical factor.
m0 <- lm(cbind(yield, tg, w100, Kconc, pvp) ~ blk + Wtr,
         data = da)

# Separated anovas.
summary.aov(m0)

# Multivariate anova.
Anova(m0)
Anova(m0, test = "Roy")

#-----------------------------------------------------------------------
# Residuals.

# Raw residuals.
r <- residuals(m0)
cor(r)

# Change the 3rd color of the palette used for the ellipses.
oldpal <- palette()
palette(c("black", "blue", "red"))
scatterplotMatrix(r,
                  gap = 0,
                  smooth = FALSE,
                  reg.line = FALSE,
                  ellipse = TRUE,
                  diagonal = "qqplot")
palette(oldpal)

corrplot(cor(r),
         type = "upper",
         tl.pos = "d",
         outline = TRUE,
         method = "ellipse")

#-----------------------------------------------------------------------

# `wtr` is a 3 level numerical factor.
m1 <- update(m0, . ~ blk + wtr + I(wtr^2))

# Separated anovas.
summary.aov(m1)

# Multivariate anova.
anova(m1) # Sequential (more appropriate for the current case).
Anova(m1) # Marginal.

anova(m1, test = "Roy") # Sequential (more appropriate ...).
Anova(m1, test = "Roy") # Marginal.

# NOTE: Univariate and Multivariate in disagreement! The way of
# representing the factors (numerical or categorical) is important. BUT
# in this case, the difference is due the type of hypothesis display
# (type I, II or III).

#-----------------------------------------------------------------------
# Performing univariate analysis.

tb <- lapply(c("yield", "tg", "w100", "Kconc", "pvp"),
             FUN = function(y) {
                 m0 <- lm(as.formula(sprintf("%s ~ blk + Wtr", y)),
                          data = da)
                 m <- LSmeans(m0,
                              effect = "Wtr")
                 g <- glht(model = m0,
                           linfct = mcp(Wtr = "Tukey"))
                 g <- cld(g, decreasing = TRUE)$mcletters$Letters
                 r <- data.frame(
                     Wtr = m$grid$Wtr,
                     means = sprintf("%0.2f %s",
                                     m$coef[, "estimate"],
                                     g[as.character(m$grid$Wtr)]))
                 names(r)[2] <- y
                 return(r)
             })

# Table of means (recursive merging to put the results together).
tbm <- Reduce(merge, x = tb)
tbm

#-----------------------------------------------------------------------
# Canonical discriminant analysis.

cd_W <- candisc(m0, term = "Wtr")
cd_W

summary(cd_W)

# The weights to compute the scores.
cd_W$coeffs.raw

# The scores.
head(cd_W$scores)

# Calculating the scores (with the centered y-variables).
Y <- m0$model[[1]]
z <- scale(Y, scale = FALSE) %*% cd_W$coeffs.raw

head(z[, 1:2])
head(cd_W$scores[, c("Can1", "Can2")])

# The loadings are the correlations between scores and responses.
cd_W$structure
cor(cbind(Y, z))[colnames(Y), colnames(z)]

# The biplot.
plot(cd_W)
```

## Analysing the subset for potassium equals 120

```{r}
#-----------------------------------------------------------------------
# Condition 2: results are similar for sets of responses.

# Subsetting the dataset.
da <- subset(swp[-74, ], pts == 120)

xtabs(~blk + Wtr, data = da)

xyplot(yield + tg + w100 + Kconc + pvp ~ wtr,
       outer = TRUE,
       scales = "free",
       type = c("p", "a"),
       data = da)

splom(~da[c(4:7, 12)],
      data = da,
      groups = Wtr,
      as.matrix = TRUE)

# `Wtr` is a 3 level categorical factor.
m0 <- lm(cbind(yield, tg, w100, Kconc, pvp) ~ blk + Wtr,
         data = da)

# Separated anovas.
summary.aov(m0)

# Multivariate anova.
Anova(m0)
Anova(m0, test = "Roy")

#-----------------------------------------------------------------------
# Residuals.

# Raw residuals.
r <- residuals(m0)
cor(r)

# Change the 3rd color of the palette used for the ellipses.
oldpal <- palette()
palette(c("black", "blue", "red"))
scatterplotMatrix(r,
                  gap = 0,
                  smooth = FALSE,
                  reg.line = FALSE,
                  ellipse = TRUE,
                  diagonal = "qqplot")
palette(oldpal)

corrplot(cor(r),
         type = "upper",
         tl.pos = "d",
         outline = TRUE,
         method = "ellipse")

#-----------------------------------------------------------------------

# `wtr` is a 3 level numerical factor.
m1 <- lm(cbind(yield, tg, w100, Kconc, pvp) ~ blk + wtr + I(wtr^2),
         data = da)

# Multivariate anova.
anova(m1) # Sequential (more appropriate for the current case).
Anova(m1) # Marginal.

anova(m1, test = "Roy") # Sequential (more appropriate ...).
Anova(m1, test = "Roy") # Marginal.

# NOTE: Univariate and Multivariate in disagreement! The way of
# representing the factors (numerical or categorical) is important. BUT
# in this case, the difference is due the type of hypothesis display
# (type I, II or III).

#-----------------------------------------------------------------------
# Performing univariate analysis.

tb <- lapply(c("yield", "tg", "w100", "Kconc", "pvp"),
             FUN = function(y) {
                 m0 <- lm(as.formula(sprintf("%s ~ blk + Wtr", y)),
                          data = da)
                 m <- LSmeans(m0,
                              effect = "Wtr")
                 g <- glht(model = m0,
                           linfct = mcp(Wtr = "Tukey"))
                 g <- cld(g, decreasing = TRUE)$mcletters$Letters
                 r <- data.frame(
                     Wtr = m$grid$Wtr,
                     means = sprintf("%0.2f %s",
                                     m$coef[, "estimate"],
                                     g[as.character(m$grid$Wtr)]))
                 names(r)[2] <- y
                 return(r)
             })

# Table of means (recursive merging to put the results together).
tbm <- Reduce(merge, x = tb)
tbm

#-----------------------------------------------------------------------
# Canonical discriminant analysis.

cd_W <- candisc(m0, term = "Wtr")
cd_W

summary(cd_W)

# The weights to compute the scores.
cd_W$coeffs.raw

# The scores.
head(cd_W$scores)

# Calculating the scores (with the centered y-variables).
Y <- m0$model[[1]]
z <- scale(Y, scale = FALSE) %*% cd_W$coeffs.raw

head(z[, 1:2])
head(cd_W$scores[, c("Can1", "Can2")])

# The loadings are the correlations between scores and responses.
cd_W$structure
cor(cbind(Y, z))[colnames(Y), colnames(z)]

# The biplot.
plot(cd_W)
```

## Analysing the entire `soybeanwp` dataset

```{r}
#----------------------------------------------------------------------
# Fitting the multivariate Gaussian linear model.

xyplot(yield + tg + w100 + Kconc + pvp ~ pts,
       groups = Wtr,
       auto.key = TRUE,
       type = c("p", "a"),
       outer = TRUE,
       as.table = TRUE,
       data = swp[-74, ],
       xlab = "Potassium",
       ylab = "Values",
       scales = list(y = "free"))

m0 <- lm(cbind(yield, tg, w100, Kconc, pvp) ~ blk + Wtr * Pts,
         data = swp[-74, ])

# MANOVA.
anova(m0)

# ANOVAs.
summary.aov(m0)

# Univariate model summaries.
# summary(m0)

# Estimated parameter.
coef(m0)

# Raw residuals.
r <- residuals(m0)
var(r)
cor(r)

# Change the 3rd color of the palette used for the ellipses.
oldpal <- palette()
palette(c("black", "blue", "red"))
scatterplotMatrix(r,
                  gap = 0,
                  smooth = FALSE,
                  reg.line = FALSE,
                  ellipse = TRUE,
                  diagonal = "qqplot")
palette(oldpal)

corrplot(cor(r),
         type = "upper",
         tl.pos = "d",
         outline = TRUE,
         method = "ellipse")

#-----------------------------------------------------------------------
# Canonical analysis.

cd_WP <- candisc(m0, term = "Wtr:Pts")
cd_WP

# Correlations between responses and canonical scores.
cd_WP$structure

# Canonical scores.
head(cd_WP$scores)

# Biplot.
plot(cd_WP)

# Merge the responses with the scores.
swps <- merge(swp, cd_WP$scores)

xyplot(-Can2 ~ -Can1 | Wtr,
       groups = Pts,
       auto.key = TRUE,
       data = swps) +
    glayer(panel.ellipse(...))

xyplot(-Can1 + -Can2 + yield + tg + w100 + Kconc ~ Pts,
       groups = Wtr,
       outer = TRUE,
       scales = list(y = "free"),
       as.table = TRUE,
       data = swps,
       type = c("p", "a"))
```

```{r eval = FALSE, include = FALSE}
#-----------------------------------------------------------------------
# ATTENTION: code in tests.

# full <- cbind(yield, tg, w100, Kconc, pvp) ~ blk + Wtr * Pts
# null <- . ~ blk + Wtr + Pts
# data <- swp[-74, ]

gugudada <- function(full, null, data) {
    # Nested models.
    m_full <- lm(full, data)
    m_null <- update(m_full, null)
    # SSP.
    S_full <- crossprod(residuals(m_full))
    S_null <- crossprod(residuals(m_null))
    S_extr <- S_null - S_full
    # Eigen decomposition.
    ei <- eigen(solve(S_full, S_extr))
    print(ei)
    scores <- as.matrix(m_full$model[1]) %*% ei$vectors[, 1:2]
    plot(scores, asp = 1)
    print(anova(update(m_full, scores[, 1] ~ .)))
    print(anova(update(m_full, scores[, 2] ~ .)))
}

# Effect of interaction.
par(mfrow = c(1, 2))
plot(cd_WP)
grid()
gugudada(full = cbind(yield, tg, w100, Kconc, pvp) ~ blk + Wtr * Pts,
         null = . ~ blk + Wtr + Pts,
         data = swp[-74, ])
grid()
layout(1)

# Effect of Wtr.
par(mfrow = c(1, 2))
plot(cd_W)
grid()
gugudada(full = cbind(yield, tg, w100, Kconc, pvp) ~ blk + Pts + Wtr,
         null = . ~ blk + Pts,
         data = swp[-74, ])
grid()
layout(1)

# ATTENTION: This requires more study to fully understand.
#-----------------------------------------------------------------------
```

# Using the `mcglm` package

TODO

# References

<!-- Insert the refereces right here! -->
<div id="refs">
</div>

# Session information

```{r}
# devtools::session_info()
Sys.time()
cbind(Sys.info())
sessionInfo()
```
