---
title: "Water retention curve parameters as function of soil management and depth"
author: |
  | Wagner H. Bonat   | Walmes M. Zeviani |
  |:-----------------:|:-----------------:|
  | `wbonat@ufpr.br`  | `walmes@ufpr.br`  |
  |  LEG/DEST/UFPR    |  LEG/DEST/UFPR    |
date: >
  62^a^ RBras & 17^o^ SEAGRO</br>
  July 24--28, 2017</br>
  UFLA, Lavras/MG
bibliography: ../slides/config/ref.bib
csl: ../slides/config/elsevier-harvard.csl
---

```{r, include = FALSE}
source("../slides/config/_setup.R")
```

# Data description and objectives

These data are measures of soil water content (m^3^ m^-3^) as a function
of the matrix potential (kPa) collected in a field under coffee
cultivation using conservation and intensive soil management.

Undisturbed soil samples were collected from a soil classified as
Distroferric Red Latosol (LVd) in a coffe crop field with 3.5 years of
implantation.

The samples were extracted from an experimental design that considered 2
levels of sampling on the crop field (inline and between lines of crop)
and depth of the soil layer (0-20, 20-40, 40-80, 80-120, 120 -160 cm),
with three replicates, making up 30 experimental units.

The tensions applied to the extracted soil samples were 0, 1, 2, 4, 6,
10, 33, 100, 500 and 1500 kPa, resulting in 300 pairs of water content
values (moisture) and matric potential (tension).

The objective of this data acquisition is to verify the effect of
management practices (made inline and between lines) on the water
retention curve of the soil as well as verify the extent of such effects
along the soil profile.

These data were used as part of the Doctor Thesis of the second author
of this course. For more information, see @zeviani2013 (in portuguese).

# Analysis with MLM

```{r, message = FALSE}
#-----------------------------------------------------------------------
# Packages.

rm(list = ls())
library(lattice)
library(latticeExtra)
library(gridExtra)
library(reshape2)
library(car)
library(candisc)
library(corrplot)
library(doBy)
library(multcomp)
```

```{r}
# Loading the data.
url <- "http://www.leg.ufpr.br/~walmes/data/ZevianiTeseCRA.txt"
wrc <- read.table(url, header = TRUE, sep = "\t")

# Position: between and on the crop lines.
# Depth: soil layers.
names(wrc) <- c("Pos", "depth", "rep", "tens", "mois")
wrc <- transform(wrc,
                 Pos = factor(Pos, labels = c("bet", "on")),
                 depth = as.integer(sub(".*-", "",
                                        as.character(wrc$depth))))
str(wrc)

# The curves.
useOuterStrips(
    xyplot(mois ~ tens | Pos + depth,
           data = wrc,
           groups = rep,
           type = "o",
           xscale.components = xscale.components.log10ticks,
           scales = list(x = list(log = 10))))

# Defines the sample unit for the water retention curve.
wrc$eu <- with(wrc, interaction(Pos, depth, rep, drop = TRUE))

# Fit the nonlinear model for each experimental unit.
model <- mois ~ Ur + (Us - Ur)/(1 + (exp(a) * (tens))^n)^(1 - 1/n)
start <- list(Ur = 0.3,
              Us = 0.6,
              a = 0.1,
              n = 1.5)

# n0 <- nls(model, data = wrc, start = start)
# summary(n0)

# Serial fit of the nonlinear model.
fits <- by(wrc,
           INDICES = wrc$eu,
           FUN = function(wrc_i) {
               n0 <- nls(model,
                         data = wrc_i,
                         start = start)
               return(cbind(wrc_i[1, c("Pos", "depth", "rep")],
                            as.data.frame(as.list(coef(n0)))))
           })

# Estimated parameters for all experimental units.
wrcp <- do.call(rbind, fits)
str(wrcp)

# The water retention parameters. Do the long and then the wide format.
wrp <- dcast(data = melt(data = wrcp,
                          id.vars = c("Pos", "depth", "rep")),
              formula = Pos + rep ~ variable + depth)
str(wrp)

# We can't apply MANOVA here unless we assume that the observations in
# depth are independent.

# Variables with capital letters denote categorical factors.
wrcp$Depth <- factor(wrcp$depth)
```
```{r, fig.height = 12}
#-----------------------------------------------------------------------
# Visualizing the data.

v <- c("Us", "Ur", "a", "n")
sp1 <- splom(~wrcp[v],
             groups = Depth,
             data = wrcp,
             auto.key = list(title = "Depth",
                             cex.title = 1,
                             columns = 3),
             par.settings = list(superpose.symbol = list(pch = 4)),
             as.matrix = TRUE)
sp2 <- splom(~wrcp[v],
             groups = Pos,
             data = wrcp,
             pch = 1,
             auto.key = list(title = "Position",
                             cex.title = 1),
             as.matrix = TRUE)

# c("Depth" = sp1, "Position" = sp2, merge.legends = TRUE)
grid.arrange(sp1, sp2, ncol = 1)
```
```{r}
#-----------------------------------------------------------------------
# Fitting the multivariate linear model.

m0 <- lm(cbind(Us, Ur, a, n) ~ Pos * Depth,
         data = wrcp)
anova(m0)
summary.aov(m0)

# splom(residuals(m0))

# Drops the interaction.
m1 <- update(m0, . ~ Pos + Depth)
anova(m0, m1)
anova(m1)

# A second order polynomial fits well.
m2 <- update(m0, . ~ Pos + depth + I(depth^2))
anova(m1, m2)
anova(m2)
# summary.aov(m2)
```

## Profile analysis

```{r}
#-----------------------------------------------------------------------
# Profile analysis.

# Inner factor data structure.
idata <- data.frame(resp = v)
Anova(m0, idata = idata, idesign = ~resp)
```

## Canonical discriminant analysis

```{r, message = FALSE}
library(candisc)

# For the effect of Position.
cd <- candisc(m1, term = "Pos")

cd
summary(cd)
plot(cd)

# For the effect of depth.
cd <- candisc(m1, term = "Depth")

cd
summary(cd)

plot(cd)
```

# Using the `mcglm` package

```{r}
# Load the package.
library(mcglm)
source("../review/functions.R")
# library(Matrix)
# packageVersion("mcglm")

# Sort the lines per subject (`loc`) followed by `cam`.
wrcp$eu <- with(wrcp, interaction(Pos, rep, drop = TRUE))
wrcp <- plyr::arrange(wrcp, eu)
head(wrcp)
```

```{r}
#-----------------------------------------------------------------------
# Covariance structures for the inner subject factor `depth`.

# Independent covariance structure.
Z0 <- mc_id(wrcp)
length(Z0)

# Distance based covariance structure.
Z_dist <- mc_dist(id = "eu",
                  time = "Depth",
                  data = wrcp,
                  method = "euclidean")
length(Z_dist)

linpred <- sapply(sprintf("%s ~ Pos * Depth",
                          colnames(m0$model[[1]])),
                  FUN = as.formula)
linpred

mc0 <- mcglm(linear_pred = linpred,
             matrix_pred = replicate(length(linpred),
                                     Z0,
                                     simplify = FALSE),
             data = wrcp)

# FIXME: this returns an error.
# mc1 <- mcglm(linear_pred = linpred,
#              matrix_pred = replicate(length(linpred),
#                                      Z_dist,
#                                      simplify = FALSE),
#              control_algorithm = list(tunning = 0.8),
#              data = wrcp)
#--------------------------------------------
# Automatic initial values selected.
# Error in .local(x, ...) :
#   internal_chm_factor: Cholesky factorization failed
# Error: $ operator is invalid for atomic vectors

#-----------------------------------------------------------------------
# Comparing models fit.

logLik.mlm(m0)
gof(mc0)
```

# References

<!-- Insert the refereces right here! -->
<div id="refs">
</div>

# Session information

```{r}
# devtools::session_info()
Sys.time()
cbind(Sys.info())
sessionInfo()
```
