---
title: "Defoliation and growth stage on the cotton production"
author: |
  | Wagner H. Bonat   | Walmes M. Zeviani |
  |:-----------------:|:-----------------:|
  | `wbonat@ufpr.br`  | `walmes@ufpr.br`  |
  |  LEG/DEST/UFPR    |  LEG/DEST/UFPR    |
date: >
  62^a^ RBras & 17^o^ SEAGRO</br>
  July 24--28, 2017</br>
  UFLA, Lavras/MG
bibliography: ../slides/config/ref.bib
csl: ../slides/config/elsevier-harvard.csl
---

```{r, include = FALSE}
source("../slides/config/_setup.R")
```

# Data description and objectives

Pests and diseases of foliage, phytotoxicity of chemicals, hail and some
mechanical injuries are the main agents of defoliation in cotton. The
purpose of this research was to study the effect of defoliation levels
at different growth stages of cotton. The study was conducted in a
greenhouseand the experimental design was completely randomized in
factorial 5 defoliation levels x 5 growth stages. The response variables
were: number of bolls produced (`nb`), number of bolls viable, total
boll weight (`wb`), plant height (`hei`) and number of plant nodes
(`nn`). The last two variables indicates the vegetative growth and the
first three indicate the plant production (or results of the
reproduction).

For more information, see @Silva2012a. @Zeviani2014 used the Gamma-Count
distribution to analyse the number of balls and found that it is a
subdispersed count variable.

# Analysis with MLM

```{r, message = FALSE}
#-----------------------------------------------------------------------
# Packages.

rm(list = ls())
library(lattice)
library(latticeExtra)
library(gridExtra)
library(car)
library(candisc)
library(doBy)
library(multcomp)
```

```{r}
# Loading the data.
url <- "http://www.leg.ufpr.br/~walmes/data/desfolha_algodao.txt"
cot <- read.table(url, header = TRUE, sep = "\t")

# Position: between and on the crop lines.
# Depth: soil layers.
str(cot)

# GS: growth stage.
# def: artificial defoliation.
# plant: plant inside plot.
# rept: treatment cellrepetition.
# nrs: number of reproductive structures (n).
# nb: number of produced balls.
# hei: plant height at hasvest.
# nn: number of plant nodes.
# wb: weight of the balls.
names(cot) <- c("GS", "def", "plant", "rept",
                "nrs", "nb", "hei", "nn", "wb")
cot <- transform(cot,
                 GS = factor(GS,
                             levels = unique(GS),
                             labels = 1:nlevels(GS)),
                 Def = factor(def))

# nrs and nb are very similar.
cot$nrs <- NULL
str(cot)
```

```{r}
#-----------------------------------------------------------------------
# Visualizing the data.

v <- c("wb", "hei", "nb", "nn")
cot <- aggregate(as.matrix(cot[, v]) ~ GS + Def + rept,
                 data = cot,
                 FUN = sum,
                 na.rm = TRUE)

combineLimits(
    useOuterStrips(
        xyplot(wb + hei + nb + nn ~ Def | GS,
               outer = TRUE,
               jitter.x = TRUE,
               as.table = TRUE,
               type = c("p", "smooth"),
               scales = list(y = "free"),
               data = cot)))
```
```{r, fig.height = 12}
sp1 <- splom(~cot[v],
             groups = GS,
             data = cot,
             auto.key = list(title = "Growth stage",
                             cex.title = 1,
                             columns = 3),
             par.settings = list(superpose.symbol = list(pch = 4)),
             as.matrix = TRUE)

sp2 <- splom(~cot[v],
             groups = Def,
             data = cot,
             auto.key = list(title = "Artificial defoliation",
                             cex.title = 1,
                             columns = 3),
             as.matrix = TRUE)

# c("Depth" = sp1, "Position" = sp2, merge.legends = TRUE)
grid.arrange(sp1, sp2, ncol = 1)
```
```{r}
#-----------------------------------------------------------------------
# Fitting the multivariate linear model.

m0 <- lm(cbind(wb, hei, nb, nn) ~ GS * Def,
         data = cot)

# Manova table.
anova(m0)

summary.aov(m0)

# Extract the raw residuals.
r <- residuals(m0)

# Checking the models assumptions on the residuals.
scatterplotMatrix(r,
                  gap = 0,
                  smooth = FALSE,
                  reg.line = FALSE,
                  ellipse = TRUE,
                  diagonal = "qqplot")

# Test on the interaction term. Preparing the hypothesis matrix.
a <- m0$assign
A <- diag(length(a))
A <- A[a == 3, ]

linearHypothesis(m0,
                 hypothesis.matrix = A,
                 test = "Pillai")
Anova(m0)

# Testing only for the `hei` and `nn` responses.
linearHypothesis(m0,
                 hypothesis.matrix = A,
                 test = "Pillai",
                 P = cbind(c(0, 1, 0, 0),
                           c(0, 0, 0, 1)))
```
```{r, message = FALSE}
library(candisc)

# For the effect of Position.
cd <- candisc(m0, term = "GS:Def")

cd
# summary(cd)

# Biplot.
plot(cd)

# Correlations between responses and canonical scores.
cd$structure

combineLimits(
    useOuterStrips(
        xyplot(Can1 + Can2 + Can3 ~ Def | GS,
               outer = TRUE,
               jitter.x = TRUE,
               as.table = TRUE,
               type = c("p", "smooth"),
               scales = list(y = "free"),
               data = cd$scores)))
```

# Using the `mcglm` package

```{r}
# Load the package.
library(mcglm)
source("../review/functions.R")
# library(Matrix)
# packageVersion("mcglm")
```

# References

<!-- Insert the refereces right here! -->
<div id="refs">
</div>

# Session information

```{r}
# devtools::session_info()
Sys.time()
cbind(Sys.info())
sessionInfo()
```
